The Name of the Title Is Hope

### 2024-12-30 paper.tex
working through text

\section{Background}
\subsection{The GPU Programming Environment(Model?)}
\john{So I have written, roughly, a zillion of these sections and here is my current philosophy. The danger is that you write something really interesting and relevant and a reviewer who is familiar with GPUs sees the title and sees the first few lines and says ``I can skip this'' and does. There are plenty of great descriptions of how GPUs work. Cite one of them. What this section needs to do is very specifically call out the interesting/detailed features of GPUs that are \emph{particularly relevant to this paper} that are not just ``here's generally how a GPU works''. For me, that's the last paragraph. The other part that might be necessary is defining the specific terminology used in the paper (e.g., workgroup vs.\ thread block vs.\ CTA). If you feel like breaking up a wall of text, which I generally think is a good idea, we can make a little table of equivalent terminology and boldface the terms we're using.} \john{Perhaps another good thing to mention that is specific to this paper is the GPU's support for synchronization. I could see having a blurb on ``there is no global inter-workgroup synchronization except an expensive kernel barrier'' and also what \emph{is} available in terms of pairwise synchronization between workgroups and what the requirements are (e.g., residence). These are more detailed (and important) GPU topics that aren't present in the higher-level summary you have below. Later in the paper, you reference ``without an inter-workgroup barrier primitive'' and don't say why, and you don't want the reader to get the wrong idea here (you don't want the reader to say ``oh, we can just add this to the language, we're done''). There is a fundamental reason to not have this primitive. Here's where you can discuss that.} \john{Thinking about the above, one way to organize this subsection that is most helpful is to \emph{show} here what kind of synchronizations are / aren't possible within the GPU programming model, in a principled way, and then these are the constraints for our implementation. We want readers to not just accept a weird set of constraints as a given but rather to understand where they come from, and why.}
Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency. Due to their specialized nature, GPUs function as co-processors to the CPU, requiring GPU programs—known as \emph{kernels} or \emph{shaders}—to be \emph{launched} or \emph{dispatched} heterogeneously by the CPU\@. Upon launch, a kernel is replicated across independent execution contexts, called threads. These threads are first organized into smaller units known as \emph{subgroups}\footnote{Also referred to as \emph{Warps}, \emph{Waves}, or \emph{SIMD-Groups}.}, which are then grouped into larger units called \emph{workgroups}\footnote{Also referred to as \emph{Cross-Thread-Arrays}, \emph{Thread-Blocks}, \emph{Blocks}, \emph{Thread-Groups}, or \emph{Groups}.}. This thread hierarchy mirrors the GPU memory hierarchy: individual threads access private, high-speed registers; threads within a workgroup share low-latency shared memory\footnote{Also referred to as \emph{Local Data Storage}, \emph{Group-Shared Memory}, (Check metal terminology).}; and all threads have access to high-bandwidth but higher-latency global memory.

@@

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Prior to the introduction of shared memory\john{pre-CUDA?}, early GPU scan implementations~\cite{Horn, Hensley, Greß, Sengupta} operated directly on global memory \john{the previous implementation strategy is PRAM-like and you should say that, it's important to a SPAA audience} and performed a single global scan where each level of depth necessitated a separate kernel launch\john{this latter phrase is not obvious! explain why}. The introduction of shared memory and dedicated compute environments resulted in the first implementations to resemble contemporary designs, combining coarse-grained inter-workgroup strategies with fine-grained intra-workgroup strategies. \emph{Work tile} partitioning was introduced to divide inputs into workgroup-consumable chunks that fit within the limited size of shared memory. However, as a natural result of this partitioning, a serial dependency is created between the work tiles. In this section, we review the three prominent global scan strategies—\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}—examining how they navigate this inter-workgroup dependency and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization in alignment with our ideal architecture goals.

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2008} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
\begin{enumerate}
  \item \textbf{Intra-Workgroup (Local) Scan:} Each workgroup performs an inclusive scan on its assigned work-tile partition of the input data. The results, including both the local scan output and the reduction of each workgroup, are written back to global memory.
  \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{the spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between workgroups; we call the result of this phase the \emph{root}.
  \item \textbf{Propagation:} The root is propagated back to each workgroup and combined with the local scan results to compute the final scan.
\end{enumerate}
Without an inter-workgroup barrier primitive, each phase is implemented as a separate kernel launch to ensure a consistent view of the reductions. During the scan operation, each element is read from and written to global memory once during the local scan and again during the propagation phase, contributing a total of $4n$ global memory operations. Although data movement is dominated by the first and third phase, Sengupta et al.'s implementation handles arbitrarily large inputs using recursion. For an input of size $n$ and a work tile of size $t$, this results in a recursive depth of $\lceil \log_t n \rceil$. At each recursive step $k$, $n/t^k$ tiles are processed, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps. Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.

\subsubsection{Reduce-then-Scan}
Due to the extreme latency of global memory access relative to compute, computationally light algorithms like \emph{scan} can benefit from redundant calculation if it reduces memory access. This insight led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated. The \emph{Reduce-Then-Scan} strategy operates in three phases:
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result into global memory.
  \item \textbf{Spine Scan:} The reductions from all workgroups are gathered and scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating the root as it writes to global memory.
\end{enumerate}
Merrill and Grimshaw~\cite{} introduced a workgroup-level \emph{raking} technique that improved the spine phase of \emph{Reduce-Then-Scan}. Instead of handling arbitrary inputs recursively, this approach determines the total workgroup occupancy $o$ and partitions the input into non-overlapping blocks, processed serially by each workgroup. By limiting the recursive depth of the global scan to two, the global memory movement for the spine scan is reduced from $O(n/t)$ to $O(3o)$. Consequently, kernel launches are limited to three, and total global memory movement is reduced to $O(3n + 3o) = O(3n)$. This approach was further refined by Ha and Han~\cite{}, who modified the upsweep behavior of the first workgroup from reduction to scan and eliminated the redundant upsweep of the last workgroup, thereby saving the equivalent of one workgroup's computation.

>>>>

\section{Background}
\subsection{The GPU Programming Environment(Model?)}
\john{So I have written, roughly, a zillion of these sections and here is my current philosophy. The danger is that you write something really interesting and relevant and a reviewer who is familiar with GPUs sees the title and sees the first few lines and says ``I can skip this'' and does. There are plenty of great descriptions of how GPUs work. Cite one of them. What this section needs to do is very specifically call out the interesting/detailed features of GPUs that are \emph{particularly relevant to this paper} that are not just ``here's generally how a GPU works''. For me, that's the last paragraph. The other part that might be necessary is defining the specific terminology used in the paper (e.g., workgroup vs.\ thread block vs.\ CTA). If you feel like breaking up a wall of text, which I generally think is a good idea, we can make a little table of equivalent terminology and boldface the terms we're using.} \john{Perhaps another good thing to mention that is specific to this paper is the GPU's support for synchronization. I could see having a blurb on ``there is no global inter-workgroup synchronization except an expensive kernel barrier'' and also what \emph{is} available in terms of pairwise synchronization between workgroups and what the requirements are (e.g., residence). These are more detailed (and important) GPU topics that aren't present in the higher-level summary you have below. Later in the paper, you reference ``without an inter-workgroup barrier primitive'' and don't say why, and you don't want the reader to get the wrong idea here (you don't want the reader to say ``oh, we can just add this to the language, we're done'', and you alternatively don't want the reader to say ``oh, we'll just use a kernel launch barrier''). There is a fundamental reason why this primitive is expensive. Here's where you can discuss that, and then assert that a global barrier (presumably a kernel launch barrier) is so expensive that we do not consider it in our implementations.} \john{Thinking about the above, one way to organize this subsection that is most helpful is to \emph{show} here what kind of synchronizations are / aren't possible within the GPU programming model, in a principled way, and then these are the constraints for our implementation. We want readers to not just accept a weird set of constraints as a given but rather to understand where they come from, and why.}
Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency. Due to their specialized nature, GPUs function as co-processors to the CPU, requiring GPU programs—known as \emph{kernels} or \emph{shaders}—to be \emph{launched} or \emph{dispatched} heterogeneously by the CPU\@. Upon launch, a kernel is replicated across independent execution contexts, called threads. These threads are first organized into smaller units known as \emph{subgroups}\footnote{Also referred to as \emph{Warps}, \emph{Waves}, or \emph{SIMD-Groups}.}, which are then grouped into larger units called \emph{workgroups}\footnote{Also referred to as \emph{Cross-Thread-Arrays}, \emph{Thread-Blocks}, \emph{Blocks}, \emph{Thread-Groups}, or \emph{Groups}.}. This thread hierarchy mirrors the GPU memory hierarchy: individual threads access private, high-speed registers; threads within a workgroup share low-latency shared memory\footnote{Also referred to as \emph{Local Data Storage}, \emph{Group-Shared Memory}, (Check metal terminology).}; and all threads have access to high-bandwidth but higher-latency global memory.

@@

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Prior to the introduction of shared memory\john{pre-CUDA?}, early GPU scan implementations~\cite{Horn, Hensley, Greß, Sengupta} operated directly on global memory \john{the previous implementation strategy is PRAM-like and you should say that, it's important to a SPAA audience} and performed a single global scan where each level of depth necessitated a separate kernel launch\john{this latter phrase is not obvious! explain why}. The introduction of shared memory \john{I think it's not just this, it's also hierarchical compute, grid$\rightarrow$workgroup$\rightarrow$thread. The important thing here is not just saying ``this is the contemporary architecture'' but also WHY it's the contemporary architecture. I like how you are focusing on the tile dependencies as the important aspect.} and dedicated compute environments resulted in the first implementations to resemble contemporary designs, combining coarse-grained inter-workgroup strategies with fine-grained intra-workgroup strategies. \emph{Work tile} partitioning was introduced to divide inputs into workgroup-consumable chunks that fit within the limited size of shared memory. However, as a natural result of this partitioning, a serial dependency is created between the work tiles. In this section, we review the three prominent global scan strategies—\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}—examining how they navigate this inter-workgroup dependency and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization in alignment with our ideal architecture goals.

\john{So if I was explaining this evolution without having read this paper, the way I would do it is by asserting that scan has a relatively low arithmetic intensity and we expect theoretically, and see experimentally, that scans are entirely limited by memory bandwidth, and in fact can run near the theoretical speed of light of the memory system. Given that, the entire goal is to make sure we organize computation in a way that allows us to use all of the theoretical memory bandwidth. Given THAT, we can talk about using the memory and compute hierarchy (shared memory, tiles, etc.) to make sure we do.}

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2008} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
\begin{enumerate}
  \item \textbf{Intra-Workgroup (Local) Scan:} Each workgroup performs an inclusive scan on its assigned work-tile partition of the input data. The results, including both the local scan output and the reduction of each workgroup, are written back to global memory.
  \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{the spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between workgroups; we call the result of this phase the \emph{root}. \john{For large scans, this stage is an insignificant amount of work compared to the first and third.}
  \item \textbf{Propagation:} The root is propagated back to each workgroup and combined with the local scan results to compute the final scan. \john{Language is a little sloppy here, since the root is not propagated back to each workgroup; instead each workgroup gets one element of the root, yes? I like that you are using and distinguishing root and spine; I just want them to be precise.}
\end{enumerate}
Without an inter-workgroup barrier primitive, each phase is implemented as a separate kernel launch to ensure a consistent view of the reductions. During the scan operation, each element is read from and written to global memory once during the local scan and again during the propagation phase, contributing a total of $4n$ global memory operations. Although data movement is dominated by the first and third phase, Sengupta et al.'s implementation handles arbitrarily large inputs using recursion. An input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$. Each recursive step $k$ processes $n/t^k$ tiles, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps. Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.

\subsubsection{Reduce-then-Scan}
Due to the extreme latency \john{I don't think this is a latency problem. GPUs tolerate long latency well. This is a bandwidth problem. We don't have enough bandwidth to keep the compute units fed.} of global memory access relative to compute, computationally light algorithms like \emph{scan} can benefit from redundant calculation if it reduces memory access. \john{I would try to use the word ``tradeoff'' here, because you'll use it again. In this case, we profit from trading additional computation for less memory bandwidth. We want to come back to ``scan is memory-bound, so reducing memory traffic, even at the cost of additional computation, is profitable''.} This insight led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated. The \emph{Reduce-Then-Scan} strategy operates in three phases:
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result into global memory.
  \item \textbf{Spine Scan:} The reductions from all workgroups are gathered and scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating the root as it writes to global memory.
\end{enumerate}
Merrill and Grimshaw~\cite{} introduced a workgroup-level \emph{raking} technique that improved the spine phase of \emph{Reduce-Then-Scan}. Instead of handling arbitrary inputs recursively, this approach determines the total workgroup occupancy $o$ and partitions the input into non-overlapping blocks, processed serially by each workgroup. \john{The important part here IMHO is changing the problem size in the second step from ``variable sized, and proportional to the problem size'' (which might require recursion) to ``constant sized, and thus guaranteed to avoid recursion''. I would identify the former as an important problem to be solved because of the potential need for further recursion, and the latter as an elegant solution to it.} By limiting the recursive depth of the global scan to two, the global memory movement for the spine scan is reduced from $O(n/t)$ to $O(3o)$. Consequently, kernel launches are limited to three, and total global memory movement is reduced to $O(3n + 3o) = O(3n)$. This approach was further refined by Ha and Han~\cite{}, who modified the upsweep behavior of the first workgroup from reduction to scan and eliminated the redundant upsweep of the last workgroup, thereby saving the equivalent of one workgroup's computation.
